---
title: "Put in your choice of a title"
author: "Your Name"
date: "Date You Finish"
output: html_document
---

# Statistics Exercise #1: Descriptive Statistics
## Ecology (Biol 404)

Data analysis in evolutionary ecology is a multi-step process. The first step is to get data into an analysis program and start to describe and visualize it. This step is critical for characterizing the data set (which will be important later for determining which kind of test to run), for error-checking, and for starting to see what your results look like.

Today, we will practice descriptive data analysis using R. Again, you should expect to experience some frustration while learning R (I still do at times), but persist. You’re learning or practicing skills (e.g., analysis, coding, logical thinking, problem-solving) within a specific programming language and software, and those skills can be useful in many contexts--not just R.

For these exercises, we will use a dataset collected by Pacific students about the distribution of poison oak on trees in a local forest. I've created "poak," a .csv file (this is the best format for importing data). 

If this data set isn't already loaded, use the code chunk below to do so. 

```{r PO data, echo = FALSE} 
#to above add eval = TRUE if data is not preloaded

url <- "https://raw.githubusercontent.com/slhalpern/PUEcology-course/refs/heads/main/PoisonOak.csv"

poak <- read_csv(file = url)
```

Keep these analyses handy—you’ll use them in the next assignment as well. **Complete the assignment individually; you are welcome to talk with each other but each person must carry out the analyses on their own.** Submit your knitted assignment through the Moodle assignment link.

## Instructions
For each task below, insert a code chunk below the task that will carry out the requested analysis/make the requested figure. Use the separate handout that provides examples of code, along with the "Getting Started in R" assignment. Remember to mark the start and end of each code chunk with the three ticks (under the Esc key), and give each chunk a unique name in the first line. I've set up a blank chunk for you in the first task that you can copy, paste, and modify for subsequent tasks, if you want. 

```{r NAME OF CHUNK}
#This is a sample chunk that you can copy, paste, and edit
#if you want. Delete these comments!

```

### Task 1
Write a code chunk (below) to make a histogram of dbh (this variable is the diameter of a tree at “breast” height, or 1.4 m off the ground, measured in cm).




Run the code chunk, and answer these questions: 
a. Does the histogram look approximately normally distributed? 


b. What are you looking at to make this judgement? 


c. If it doesn’t look normally distributed, how does it differ from normal, and how “badly”?


### Task 2  (NOTE THEY ALREADY DID THIS IN THE LAST ASSIGNMENT)
Write a code chunk below to make a box plot of dbh, grouping by Presence (this variable is whether or not the tree had poison oak growing on it) and paste it below. Answer these questions: 
Does the variance look approximately equal between the groups? 
What are you looking at to make this judgement? 
If it doesn’t look equal, how does it differ, and how “badly”?

