---
title: "Put in your choice of a title"
author: "Your Name"
date: "Date You Finish"
output: pdf_document:
  toc: yes
---

# Getting Started in R

*This exercise was modified from Grayson et al. 2022 <https://www.sciencedirect.com/science/article/pii/S0025556422000499>, and they modified it from Data Visualization: A Practical Introduction by Kieran Healy <https://socviz.co/gettingstarted.html#gettingstarted> (Ch. 2).*

Welcome! This document introduces the statistical and graphical program R. R one of the most widely used programs for data analysis, including in biology. One reason R is great is that it's freeware, so anyone can use it. It's also set up to facilitate reproducibility because all analyses are done using code. This also makes is harder because you don't just point and click buttons. R Studio provides a nicer interface for working with R, including a space to write scripts (saved code that you can revisit easily). We're going to make it even easier by using R Markdown, which combines code with text where you can explain what you did and what it means--so like a lab notebook for data analysis. 

## This Document is an RMarkdown File

R Markdown combines sections of text (like this) with sections of code (termed "code chunks") that do things like calculations, figure generation, etc. The code chunks get executed in R, through the R Studio interface, which you're working in now. When you finish, you can produce a final "knitted" document that combines text and output from code in a formatted style that is ready to share with different audiences--or to turn in as an assignment! *(Note: Submit your knitted assignments through Moodle assignment links.)*

By working through this document, you'll learn some basics of R, R Studio, and R Markdown. *Make sure to type your answers to questions/completed tasks directly into this document, so they're included in the final assignment you turn in.* 

At the very top of this R Markdown file is a section of *metadata*, or information about what the file is and what it does. The metadata is delimited by three dashes at the start and another three at the end. If you create your own new R Markdown file in the future, the program automatically sets up the metadata, though you can always edit them later. 

**TASK: Change the title, author, and date in the metadata section to your values!** For now, keep the `output` line as it is; it determines the format when you create an exported document (terming "knitting"). **ADVANCED option (if you have experience with R/R Markdown): Code the date to automatically update to today's date.**

You can type in a Markdown file just like you type text in a word-processing document. Give it a try to answer these questions--place the cursor after the asterisk following "Your answer", and just type. 

**QUESTION: How much have you done coding or used the R programming language before?**

*Your answer* 

**QUESTION: Describe your experience and comfort level with coding or R in the space below.** Note: If you have experience, please tackle the advanced options as you work through this handout. If you are new to R, please consult us and your classmates with questions along the way as we develop our skills together!

*Your answer* 

## Code chunks

In R Markdown, **code chunks** are specially delimited sections of the file for code, which runs in R. The green arrow on the chunk's right will run the code within the chunk in the R console (bottom left corner of the R Studio frame--see image in the Intro to R handout). You should see the output both in the console and below the code chunk. A few points about code chunks: 

* Code chunks start and end with three **`** marks (not single quotation marks, but the bottom of the tilda button under the Esc key). 
* Everything in between the start and end marks is R code that will be executed in the console of R Studio. 
* Code chunks also must start by identifying the programming language, always r for us, in curly brackets (see below); in those brackets, you can also name the chunk or set other options, though it's not required. If you name chunks, each must have a unique name. 
* Use hashtags to comment within code. This is generally best practice for annotating/explaining code, but it should be short. For more extensive comments or commentary, use the text sections of the Markdown file.
* RStudio usually does some color coding that can be helpful--for example, comments are a different color than code,and some commands (and values) are a different color too.
* You will gain experience by modifying the code, more and more so through the next few weeks. Look for instructions in the text, and associated with some # in the gray code chunks. 

```{r chunk1}
# This is your first code chunk
# You can include text within code chunks, but you need a hashtag before it--this tells R to ignore it (i.e., it's not code). Hashtagged lines are called comments.

# R can do math: (remember, this is a comment because of the #)
2+2

# But many R commands use functions, like these simple ones:
sqrt(10)

print("R code is cool")

# When you click the run arrow, the output should appear below the chunk and in the RStudio Console

```

You can add your own chunk by moving the cursor to a blank line and choosing Code \> Insert Chunk from the RStudio menu, or clicking the green square with the white C and the + sign in the bar above then selecting R. When you do either of these, an empty chunk will appear in your file. You can also type 3 **`**, add your code, then close with another 3 **`**.

**TASK: Create your own code chunk in the blank lines below.** Write some example code similar to the first chunk above (or any code that know that you would like to include).




## EXPORTING your Markdown file

When you click the **Knit** button at the top (the blue ball of yarn icon), it will generate a document that includes both content as well as the output of any embedded R code chunks within the document (unless you exclude it from the document, like the next setup code chunk below). Save your R Markdown (Rmd) file (click the disc icon in the top left, or Ctrl S) and try it out now! Note that you'll knit again at the end when you finish to generate your final output to turn in.

**ADVANCED options**: Learn about customizing the appearance of code chunks in Markdown: <https://r4ds.had.co.nz/r-markdown.html#code-chunks>

The open source book "R for Data Science" is the fundamental R skills book by Hadley Wickham (Chief Scientist at RStudio) and Garrett Grolemund (Principal Data Science Educator at RStudio) <https://r4ds.had.co.nz/> Check out section 27.4.2 and try out some of the display options in your practice chunk (*again, optional for those who are experienced using R*)



## Load Libraries

Packages are sets of code (and sometimes data) that others have written to make it easier to carry out certain analyses in R--like making a figure or calculating summary statistics. They are kind of like apps on a device. To use them, you first have to install them (like downloading an app), which adds them to your directory, or library, in R. Then you have to load them (like opening the app). You only have to install packages once; after that, you only need to load them from your library when you start a session in R. You can install packages from the "Packages" tab in the "Files" frame in the bottom right corner of R Studio; you can also type the command `install("name of package")` in the console. (Note the quotation marks!)

To use installed packages, we use the command `library(name of package)`. Without this step, R will not be able to find the functions contained in these packages. The best practice is to load (using library commands) all the packages we need in a first code chunk at the top of the R Markdown document, or at the start of an R script if you're working there. 

In Posit Cloud, I install the packages you need, but you still need to run the code to load the libraries. Today we will be using the tidyverse family of functions, which includes `ggplot`. The chunk below loads these libraries.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# opts_chunk$set() is a function that will affect all code chunks in the report.
# echo = TRUE displays both the code and results in the knitted file.
# The global chunk options become the defaults for the rest of the document. 
# If you want a particular chunk to have a different appearance you can specify a different option within that chunk (as we've done here with include=FALSE in the header)

# If you are working in R Studio Cloud, these packages have already been installed in this workspace. If you are using R Studio Desktop, you would need to first check that these packages are locally installed. If they are not, you can delete the hashtags below to run the install code. 

#install.packages("tidyverse")
#install.packages("socviz")
#install.packages("tinytex")
#install.packages("knitr")

library(tidyverse)
library(socviz)
library(tinytex)
library(knitr)
```

## A few fundmental things to know about R

### Everything is an object

The R coding language is based on objects. These can take many forms, but often include data that can be numbers, words/characters, lists, or even data tables. They can also be output from analyses (e.g., a figure or results from a statistical test). You can directly type objects into R, import them from other files such as a spreadsheet, or generate them using code. Assign a name to objects using a left-facing "arrow" as in: Name <- data. After you name an object, you can refer to it later, and R will use whatever is to the right of the arrow when carrying out a function. *Hint: Make names short, and avoid any R code terms such as mean* Note: You can see the active objects in the environment window on the top right of your screen

**TASK:** Run the code chunk below (click the right green arrow button). Add annotations next to the empty hashtags, as demonstrated in the first line of code, to indicate what the code is doing: *NOTE THAT THIS TASK MAY REQUIRE MORE BACKGROUND--PROVIDE HERE OR IN READING??*

```{r create_objects}
c(1, 2, 3, 1, 3, 5, 25)
# example of annotating the code: 
# c() combines values into a vector, an example of an R object
# These should print below the chunk when it runs

my_numbers <- c(1, 2, 3, 1, 3, 5, 25)
# 

your_numbers <- c(5, 31, 71, 1, 3, 21, 6)
# 
```

**TASK:** Run the code chunk below (click the right green arrow button) and annotate the code with what it does. 

```{r print_object}
my_numbers

# 

```


### Do things in R using functions

A function's arguments are the things the function needs to know in order to do something.

```{r this_will_error, eval = FALSE}
# functions need arguments
mean()

# The Markdown file will not knit with an error in the code, but we wanted to demonstrate 
# getting an error. Thus this chunk has eval = FALSE in the header to display the code but suppress running the chunk or displaying the output
```

**QUESTION:** Why do you get an error? How do you find out what arguments the mean function needs? (You will fix it in the next code chunk) *Hint: run this chunk and look at the output in the console pane or below the chunk--it will give you information about the error.*

*Your answer:*


**TASK:** Follow the example below to add code to the chunk that will calculate the mean of the object your_numbers

```{r mean function_correct}
mean(x = my_numbers)

# While the function arguments have names that are used internally, (here x), 
# you don’t strictly need to specify the name for the function to work, as the next line shows.

mean(my_numbers)

# below add code to calculate the mean of your_numbers



```

Let's try another function: `summary()`

```{r summary_example}
summary(my_numbers)

# You can also assign results to an object
my_summary <- summary(my_numbers)

# Then you can "call" the summary by just typing the name:
my_summary
```

**QUESTION:** Run the code in the chunk aboven. Based on the output, what does this function (`summary`) do?

*Your answer*


### Functions come in libraries

Functions from libraries are "pre-coded" to perform tasks without writing commands from scratch. 

**TASK:** Run the code, then add annotations next to the hashtags in the code chunk to note what each of these functions doing:

```{r more_functions}

table(my_numbers)
#

sd(my_numbers)
#

```


### If you're not sure what an object is, ask for its class

The code chunk below gives some examples for doing this. Run it to see what the output looks like. 

```{r getclass}
class(my_numbers)
class(my_summary)
class(summary)
```

Objects can change class based on the inputs:

```{r change_class}
my_new_vector <- c(my_numbers, "Apple")

```

**TASK: Create your own code chunk that answers the questions below**



**QUESTION: How do you look at the information in your new object `my_new_vector`?** Hint: review the code chunk called print_object.) 

*Your answer:* 

**QUESTION: What data class does your new object contain?**

*Your answer:*


## Working with datasets

Some libraries have data files associated with the package of functions. The socviz library has a small dataframe called "titanic." Let's look at it:

```{r titanic1}
titanic
class(titanic)
```

You can call just a portion of the data using the $ to designate a column. Let's try it: 

```{r titanic2}
titanic$percent

```

**TASK: How would you select the sample size (n) column in the titanic dataset?** Create your own code chunk below to return the values in the "n" column. Remember to include the {r} line.




The tidyverse (a powerful package used in biology and data science) uses an augmented version of a data frame called a tibble. Tibbles have some features that can become important. For example, on its own it will never change the name of variable, or the type of input (e.g., from factor to character). Let's make R treat the previous dataframe as a tibble:

```{r titanic3}
titanic_tb <- as_tibble(titanic) #"as_tibble" makes R treat the data frame as a tibble
titanic_tb
```

**QUESTION:** What differences (if any) do you see in the data frame vs. the tibble in this example?

*Your answer:*


### Introduction to graphing with R

R is does an amazing job with figures! We'll focus on using the function `ggplot` to make figures. 

Here, I just want you to see making a graph in R. Follow the code I provide. In later assignments, I'll have you practice making a figure like this one with a new data set and show you ways to build and customize your own figures.

```{r intro ggplot}

ggplot(data = titanic_tb, aes(x = sex, y = percent)) + #this line identifies the data set, x, and y variables
  geom_boxplot() #this line says how to graph the data--the "geom" identifies the type of graph to make

# Note for the first line of code: R assumes x variable is first, then y. 
# You can explicitly identify them in the code as done here
# You can also just list things in the standard R order. The less-explicit first line would then be:
# ggplot(titanic_tb, aes(sex, percent))
# These two formats are equivalent.
```


## This is important! Be patient with R, and with yourself

Here are three very specific things to watch out for when writing R code:

-   Make sure parentheses are balanced and that every opening "`(`" has a corresponding closing "`)`".
-   Make sure you complete your expressions. If you think you have completed typing your code, but instead of seeing the `>` command prompt at the console you see the `+` character instead, that may mean R thinks you haven't written a complete expression yet. You can hit `Esc` or `Ctrl-C` to force your way back to the console and try typing your code again.
-   In ggplot specifically, we will build up plots a piece at a time by adding expressions to one another. *When doing this, make sure your `+` character goes at the end of the line, and not the beginning.* That is, write this:

```{r ggplot1}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
    geom_point()
```

and not this:

```{r ggplot2, echo=TRUE, eval=FALSE}
ggplot(data = mpg, aes(x = displ, y = hwy)) 
+ geom_point()

# What do you need to fix here? Change it to be correct!
# Note: Our header includes echo=TRUE [show the code], eval=FALSE [don't run the code] 
# for the knitted output document since we know there is an error. Leave 
# this header so your fixed code prints in the output file, but not a duplicate of the graph
```

Congrtulations! You've finished your introduction to R assignment!