---
title: "Put in your choice of a title"
author: "Your Name"
date: "Date You Finish"
output: pdf_document:
  toc: yes
---

# Intro to importing, summarizing, & plotting data

Welcome! This document introduces the basics of working with data in R--importing data sets, summarizing variables, and plotting data (in other words--making graphs). You will use these tools frequently, so keep your work handy as a reference in the future.

## Load Libraries

Here are the packages to load for this assigment. Run this code before anything else! If you need to install the packages, remove the # in front of the first 4 lines of code.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("tidyverse")
#install.packages("socviz")
#install.packages("tinytex")
#install.packages("knitr")

library(tidyverse)
library(socviz)
library(tinytex)
library(knitr)
```

## Get data into R

To practice more with R and graphing, we are going to use a dataset collected by Pacific students about caterpillars eating kale plants. We'll call the file kale. I've saved a .csv file (this is the best format for importing data), and we'll import it from github using the code below. Later, I'll show you how to import your own data sets. 

```{r REU data, echo = TRUE} 
#to above add eval = FALSE if data preloaded

url <- "https://raw.githubusercontent.com/slhalpern/PUEcology-course/refs/heads/main/REU_2021.csv"

kale <- read_csv(file = url)
```


## Summarizing data in various ways

Let's start by practicing something you've done before, now using the data you just imported. We'll look at the variable RGR, or relative growth rate, a measure of how fast an organism grows measured as g/g/day. 

**Task: I've started a code chunk below. Add code to it that will summarize the variable RGR.**  *Hints: Return to the titanic data from the previous assignment to remember how to select a specific column. Make sure you have the names of the columns correct, lowercase and uppercase characters matter in variable names.*

```{r intro_kale}

kale #remind yourself what this code does! It's OK to run the chunk to see what happens.

#delete this comment and add code to summarize RGR here!

```

**QUESTION:** What is the average RGR? What is the range of values for RGR?

*Your answer:* 


## Summarizing data by groups
In biology, we're often interested in calculating central tendency and spread variables by groups--e.g., comparing a treatment and a control group. We can do that in the `tidyverse` by using `group_by`. In the kale data set, groups include:

* Temporal_Rep (week of the trial)
* Mvmt_Trt (whether larvae could move between plants), and 
* Fertilizer_Trt (whether plants were fertilized or not). 

The code below shows how to summarize a variable by group.

```{r summarizing kale by group}
kale |> #|> tells R to take the output of this line and pass it to the next line of code--it's called a pipe
  group_by(Fertilizer_Trt) |> # this groups data by fertilizer treatment. Note the pipe again!
  summarize(avg.RGR = mean(RGR, na.rm = TRUE)) #na.rm = TRUE tells R to ignore any missing values, which are coded as NA in the data table
```

**Question:** Is RGR faster or slower in fertilized kale plants compared to unfertilized ones?

*Your answer:*


**Task** Revise the code in the chunk below to calculate average RGR for one of the other groups. Run the chunk and answer the questions below. 


```{r summarizing kale by group2}
kale |> #|> tells R to take the output of this line and pass it to the next line of code--it's called a pipe
  group_by(Fertilizer_Trt) |> # this groups data by fertilizer treatment. Note the pipe again!
  summarize(avg.RGR = mean(RGR, na.rm = TRUE)) #na.rm = TRUE tells R to ignore any missing values, which are coded as NA in the data table
```


**Questions:** 

What grouping variable did you choose? *Your answer:*

Is which group is your variable larger? *Your answer:*

Does the effect of the group you chose on your variable make sense ? Briefly (in a sentence or so) explain your reasoning. *Your answer:*



We might want to calculate more than just the mean for groups. For example, we might want to know how many observations are in each group (`length` counts this), the standard deviation for each group (`sd`), the variance for each group (`var`), the smallest (`min`) and largest (`max`) value for each group, or the median for each group (`median`). 

**Task: Revise the code chunk below to calculate these different variables.** Start where the comment says "fill in new code here." Hint: If you want to see them all in one table, put a comma in between the last 2 parentheses in the `summarize` line of code, hit return, add the next variable (be sure to name it before the = sign!), and repeat. I have done this for you in the code chunk below--the final parentheses is at the end. *Note: Do NOT use na.rm = TRUE for length*

```{r summarizing kale by group3}
kale |> 
  group_by(Fertilizer_Trt) |> 
  summarize(avg.RGR = mean(RGR, na.rm = TRUE),
            #fill in new code here
  ) #this parentheses closes the "summarize" command--without it you'll run into an error
```


## Make a figure

We will now use the kale data to make two figures using `ggplot`. I will walk you through the first one, then you will make subsequent ones more independently. Keep this file handy as a model for code you'll need later!

We'll start by making a histogram of the rgr data. Run this chunk (again, click the green arrow in the top right corner of the chunk). 

```{r histogram}

ggplot(data = kale, aes(x = RGR))
# Reminders: 
# Put the name of your data file after data =
# aes (short for aesthetic) tells R about the variables you'll use in the plot--here it's just one because it's a histogram
# information for aes always goes inside parentheses; if there is more than one piece of information, they are separated by commas. See the Titanic examples in the previous assignment.
# Watch your parentheses! One of the most common errors is missing closing one! 
```

Ooops! Why is the graph blank? 

We need to add a *geom*, which tells R what kind of graph to make! 

**TASK** Add `geom_histogram()` to the code below using the plus sign at the *end* of the first line (with the geom on the 2nd line). *Note: Often you will see ggplot code written with each argument on a different line to make it easier to follow; again, each line ENDS with a + sign, which tells R to add the next line to the figure being built.*

```{r histogram add geom}

ggplot(data = kale, aes(x = RGR))


```

**QUESTION:** Describe the histogram you created--it's shape, distribution of the data, etc.

*Your answer:*

**ADVANCED:** If you have used R and ggplot before, or just want to explore it, try changing the bin width or number of bins, the orientation (horizontal vs. vertical), or something else you want to play with. <https://ggplot2.tidyverse.org/reference/geom_histogram.html> has information on the code. 


**TASK: make a boxplot, using geom_boxplot**
Boxplots are some of the most useful figures in ecology because they provide a lot of information about the data. Practice them by making a boxplot for rgr separated by the grouping variable of fertilizer treatment. Refer to the example with the Titanic data from the previous assignmnet and write (or edit!) code to make the boxplot.  

**ADVANCED:** If you have used R and ggplot before, or just want to explore it, try changing the color, scale, or some other display feature of the boxplot. See this guide for examples of changes you can make to boxplots, and the code to do so: <https://www.sthda.com/english/wiki/ggplot2-box-plot-quick-start-guide-r-software-and-data-visualization>

```{r boxplot}

# Delete this and put your code here! You can copy and revise your prior code chunk if you want.


```

**QUESTION:** Describe the boxplot you created, including information about both the central tendency and the spread. 

*Your answer:*


**Congrats!** You have written code and produced figures! Next we'll make more exploratory figures, and start to carry out statistical analyses. 


